Memory for television episodes preserves event content while introducing new across-event similarities

### 2020-02-13 paper/main.tex
corrected stats in text, fixed reference to discussion section, other minor wording changes

We can also apply two list-learning-native analyses that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed video events were clustered heavily in participants' recall event sequences (mean: 0.767, SEM: 0.029), and that participants with higher temporal clustering scores tended to perform better according to both \cite{ChenEtal17}'s hand-annotated memory scores (Pearson's $r(15) = 0.62,~p = 0.008$) and our model's estimate (Pearson's $r(15) = 0.54,~p = 0.024$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar video events together (mean: 0.787, SEM: 0.018), and that semantic clustering score was also related to both hand-annotated  (Pearson's $r(15) = 0.65,~p = 0.004$) and model-derived (Pearson's $r(15) = 0.63,~p = 0.007$) memory performance.
% NOTE: 
% Semantic and temporal clustering are both positively correlated with n_events -- r=0.49; p=0.046 and r=0.61; p=0.009, respectively.  
% - Paxton

@@

Statistical models of memory studies often treat recall success as binary (i.e., an item either was or was not recalled), or occasionally categorical \citep[e.g., to distinguish familiarity from recollection;][]{YoneEtal02}.  Such approaches are tenable in classical list-learning or recognition memory paradigms, as the presented stimuli tend to be very simple (e.g., a sequence of individual words or items).  However, the feature-rich content of a naturalistic experiences may later be described with many, highly variable levels of success.  Our framework produces a content-based model of individual stimulus and recall events by projecting the dynamic content of the video and participants' recalls into a shared topic space.  This allows for direct, quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel, \textit{continuous} metrics for analyzing naturalistic memory:~\textit{precision} and \textit{distinctiveness}.  We define precision as the ``completeness" of recall, or how fully the presented content was recapitulated in memory.  Under our framework, we quantify this for a given recall event as the correlation between the topic proportions of the recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  A second novel metric we introduce here is \textit{distinctiveness}, which we define as the ``specificity" of recall, or how unique the description of a given section of content was, compared to descriptions for other sections of content.  We quantify this for each recall event as 1 minus the average correlation between the given recall event and all other recall events not matched to the same video event.  In addition to individual events, one may also use these metrics to describe each participant's overall performance (i.e., by averaging across a participant's event-wise precision or distinctiveness scores).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that, across participants, a higher precision score is correlated to both hand-annotated memory performance (Pearson's $r(15) = 0.56, p = 0.021$) and the number of video events successfully remembered, as determined by our model (Pearson's $r(15) = 0.88, p < 0.001$).  We also hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  We find that this distinctiveness score is related to our model's estimated number of recalled events (Pearson's $r(15) = 0.55, p = 0.021$), and while we do not find distinctiveness to be related to hand-annotated memory performance (Pearson's $r(15) = 0.28, p = 0.275$), this is not entirely surprising given how the hand-annotated memory scores were computed (see \textit{Methods}).

@@

In extending classical free recall analyses to our naturalistic memory framework, we recovered two patterns of recall dynamics central to list-learning studies: a heightened probability of initiating recall with the first presented ``item" (in our case, video events; Fig.~\ref{fig:list-learning}A) and a strong bias toward transitioning from recalling a given event to recalling the one immediately following it (Fig.~\ref{fig:list-learning}B).  However, equally noteworthy are the typical free recall results \textit{not} recovered in these analyses, as each highlights a fundamental difference between the list-learning paradigm and naturalistic memory paradigms like the one employed in the present study.  The most noticeable departure from hallmark free recall dynamics in these findings is the apparent lack of a serial position effect in Figure \ref{fig:list-learning}C, which instead shows greater and lesser recall probabilities for events distributed across the video.  Stimuli in free recall experiments most often comprise lists of simple, common words, presented to participants in a random order. \citep[In fact, numerous word pools have been developed based on these criteria; e.g.,][]{FrieEtal82}.  These stimulus qualities enable two assumptions that are central to word list analyses, but frequently do not hold for real-world experiences.  First, researchers conducting free recall studies may assume that the content at each presentation index is essentially equal, and does not possess attributes that would render it, on average, more or less memorable than others.  Such is rarely the case with real-world experiences or experiments meant to approximate them, and the effects of both intrinsic and observer-dependent factors on stimulus memorability are well established \citep[for review see][]{ChunTurk07, ByliEtal15, TyngEtal17}.  Second, the random ordering of list items ensures that (across participants, on average) there is no relationship between the thematic similarity of individual stimuli and their presentation positions---in other words, two successively presented items are no more likely to be highly semantically similar than they are to be high dissimilar.  In most cases, the exact opposite is true of real-world episodes.  Our internal thoughts, our actions, and the physical state of the world around us all tend to follow a direct, causal progression.  As a result, each moment of our experience tends to be inherently more similar to surrounding moments than to those in the distant past or future.  Memory literature has termed this strong temporal autocorrelation ``context," and in various media that depict real-world events (e.g., movies or written stories), we recognize it as a \textit{narrative structure}.  While a random word list (by definition) has no such structure, the logical progression between ideas and actions in a naturalistic stimulus prompts the rememberer to recount presented events in order, starting with the beginning.  This tendency is reflected in our findings' second departure from typical free recall dynamics: a lack of increased probability of first recall for end-of-sequence events (Fig.~\ref{fig:list-learning}A).

@@

% \bibliography{../../CDL-bibliography/memlab}
\bibliography{CDL-bibliography/memlab}

>>>>

We can also apply two list-learning-native analyses that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed video events were clustered heavily in participants' recall event sequences (mean: 0.767, SEM: 0.029), and that participants with higher temporal clustering scores tended to perform better according to both \cite{ChenEtal17}'s hand-annotated memory scores (Pearson's $r(15) = 0.62,~p = 0.008$) and our model's estimate (Pearson's $r(15) = 0.54,~p = 0.024$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar video events together (mean: 0.787, SEM: 0.018), and that semantic clustering score was also related to both hand-annotated  (Pearson's $r(15) = 0.65,~p = 0.004$) and model-derived (Pearson's $r(15) = 0.63,~p = 0.007$) memory performance.
% NOTE: 
% Temporal and semantic clustering are both positively correlated with n_events -- r=0.49; p=0.046 and r=0.61; p=0.009, respectively.  
% - Paxton

@@

Statistical models of memory studies often treat recall success as binary (i.e., an item either was or was not recalled), or occasionally categorical \citep[e.g., to distinguish familiarity from recollection;][]{YoneEtal02}.  Such approaches are tenable in classical list-learning or recognition memory paradigms, as the presented stimuli tend to be very simple (e.g., a sequence of individual words or items).  However, the feature-rich content of a naturalistic experiences may later be described with many, highly variable levels of success.  Our framework produces a content-based model of individual stimulus and recall events by projecting the dynamic content of the video and participants' recalls into a shared topic space.  This allows for direct, quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel, \textit{continuous} metrics for analyzing naturalistic memory:~\textit{precision} and \textit{distinctiveness}.  We define precision as the ``completeness" of recall, or how fully the presented content was recapitulated in memory.  Under our framework, we quantify this for a given recall event as the correlation between the topic proportions of the recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  A second novel metric we introduce here is \textit{distinctiveness}, which we define as the ``specificity" of recall, or how unique the description of a given section of content was, compared to descriptions for other sections of content.  We quantify this for each recall event as 1 minus the average correlation between the given recall event and all other recall events not matched to the same video event.  In addition to individual events, one may also use these metrics to describe each participant's overall performance (i.e., by averaging across a participant's event-wise precision or distinctiveness scores).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that, across participants, a higher precision score is correlated to both hand-annotated memory performance (Pearson's $r(15) = 0.60, p = 0.010$) and the number of video events successfully remembered, as determined by our model (Pearson's $r(15) = 0.90, p < 0.001$).  We also hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  We find that this distinctiveness score is related to our model's estimated number of recalled events (Pearson's $r(15) = 0.55, p = 0.021$), and while we do not find distinctiveness to be related to hand-annotated memory performance (Pearson's $r(15) = 0.28, p = 0.275$), this is not entirely surprising given how the hand-annotated memory scores were computed (see \textit{Discussion}).

@@

In extending classical free recall analyses to our naturalistic memory framework, we recovered two patterns of recall dynamics central to list-learning studies: a heightened probability of initiating recall with the first presented ``item" (in our case, video events; Fig.~\ref{fig:list-learning}A) and a strong bias toward transitioning from recalling a given event to recalling the one immediately following it (Fig.~\ref{fig:list-learning}B).  However, equally noteworthy are the typical free recall results \textit{not} recovered in these analyses, as each highlights a fundamental difference between the list-learning paradigm and naturalistic memory paradigms like the one employed in the present study.  The most noticeable departure from hallmark free recall dynamics in these findings is the apparent lack of a serial position effect in Figure \ref{fig:list-learning}C, which instead shows greater and lesser recall probabilities for events distributed across the video.  Stimuli in free recall experiments most often comprise lists of simple, common words, presented to participants in a random order. \citep[In fact, numerous word pools have been developed based on these criteria; e.g.,][]{FrieEtal82}.  These stimulus qualities enable two assumptions that are central to word list analyses, but frequently do not hold for real-world experiences.  First, researchers conducting list-learning experiments may assume that the content at each presentation index is essentially equal, and does not possess attributes that would render it, on average, more or less memorable than others.  Such is rarely the case with real-world experiences or experiments meant to approximate them, and the effects of both intrinsic and observer-dependent factors on stimulus memorability are well established \citep[for review see][]{ChunTurk07, ByliEtal15, TyngEtal17}.  Second, the random ordering of list items ensures that (across participants, on average) there is no relationship between the thematic similarity of individual stimuli and their presentation positions---in other words, two successively presented items are no more likely to be highly semantically similar than they are to be high dissimilar.  In most cases, the exact opposite is true of real-world episodes.  Our internal thoughts, our actions, and the physical state of the world around us all tend to follow a direct, causal progression.  As a result, each moment of our experience tends to be inherently more similar to surrounding moments than to those in the distant past or future.  Memory literature has termed this strong temporal autocorrelation ``context," and in various media that depict real-world events (e.g., movies or written stories), we recognize it as a \textit{narrative structure}.  While a random word list (by definition) has no such structure, the logical progression between ideas and actions in a naturalistic stimulus prompts the rememberer to recount presented events in order, starting with the beginning.  This tendency is reflected in our findings' second departure from typical free recall dynamics: a lack of increased probability of first recall for end-of-sequence events (Fig.~\ref{fig:list-learning}A).

@@

\bibliography{../../CDL-bibliography/memlab}
% \bibliography{CDL-bibliography/memlab}
